---
title: "chapter-one-exercise"
author: "Anyanwu Chinedu"
output: pdf_document
---
##### Chapter one exercise

##### Package(s) & dataset(s) used in this exercise includes:

* tidyverse
* mpg data set

###### ``mpg dataset`` represent the Fuel economy data from the 1999 to 2008 for 38 popular models of cars

***Question: Import the tidyverse package***

###### Answer:
```{r import the tidyverse library, message=FALSE, warning=FALSE}
library(tidyverse)
```

***Question: Run the ggplot(data=mpg). What do you see?***


```{r, fig.height=2, fig.width=3}
ggplot(data=mpg)
```

###### Answer:
**The visual shows a blank output. This is because no mapping variables were specified**


***Question: What does the ``drv`` variable describe?***

###### Answer:
**This is the type of drive train, comprising f=front-wheel r=rear wheel drive,4=4wd. This is according to the documentation on executing ``?mpg``**


***Question: Make a scatterplot of ``hwy`` versus ``cyl``***


###### Answer:
```{r, fig.height=2, fig.width=3}
ggplot(data = mpg)+
  geom_point(mapping = aes(x = hwy, y = cyl))
```


***Question: What happens if you make a scatterplot of ``class`` versus ``drv``? Why is the plot not useful?***


```{r, fig.height=2, fig.width=3}
ggplot(data=mpg)+
  geom_point(mapping=aes(x = class, y=drv))
```


###### Answer:
**Both ``dr`` & ``class`` are categorical variables taking small number of values. Hence the scatterplot is not useful**


***Question: What's wrong with this code? Why are the points not blue?***

```{r, fig.height=2, fig.width=3}
ggplot(data=mpg)+
  geom_point(mapping = aes(x = displ, y = hwy, color ="blue"))
```


###### Answer:
**The ``variable`` to the ``color`` argument as used here must be a known variable to the data set. However, to make the points blue, the argument must be specified outside of ``aes() as shown below``**

```{r, fig.height=2, fig.width=3}
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy), color="blue")
```


***Question: Which variables in mpg are categorical? Which variables are continuous?***


###### Answer:
**Categorical variables are discrete variables that take limited amount of values(eg - yes/no, dead/alive, male/female/binary, etc.), while continuous variables take fairly large number of values(eg - cities in Eastern Nigeria, species of Penguins etc.)**

**By running the command ``View(data=mpg)``, we can discover that ``class,drv, fl, hwy, cyl`` are all 
categorical variables while the variables ``manufacturer, model, displ, year, cty, hwy`` are continuous**


***Question: What happens if you facet on a continuous variable?***


###### Answer:
**The visualization results in a more sub visuals with each continuous variable converting to a discrete variable**


***Question: What do empty cells in a plot with ``facet_grid(drv ~ cyl) mean? How do they relate to this plot?***

```{r, fig.show='hide'}
ggplot(data = mpg)+
  geom_point(mapping = aes(x = drv, y =cyl))
```


###### Answer:
**Empty cells in such as plot as this show points of no observations**


***Question: What geom would you use to draw a line a chat? A boxpplox? A histogram? An area chart?***

**``geom_line(), geom_boxplot(), geom_histogram and geom_area``**

***Will these two graphs look different? why/why not?***


```{r fig.show='hide', message=FALSE, warning=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy))+
  geom_point()+
  geom_smooth()
```


```{r, fig.show='hide', message=FALSE, warning=FALSE}
ggplot()+
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy))+
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```


###### Answer:
**The both plots will not look different. In the first plot, the mapping and variables were made global so each of the ``geoms`` has access to the same data. However, in the second plot, the same mapping and variables were made local to the respective ``geoms``**


***Question: Re-create the R code necessary to generate the following graphs***


```{r, message=FALSE, warning=FALSE, echo=FALSE, fig.height=2, fig.width=3, fig.align='center'}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy))+
  geom_point()+
  geom_smooth(se = FALSE,color = "blue")

```

```{r, warning=FALSE, echo=FALSE, message=FALSE,fig.align='center', fig.width=3, fig.height=2}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy))+
  geom_point()+
geom_smooth(se = FALSE, mapping = aes(x = displ, y = hwy, group_by =drv), show.legend = FALSE)

```

```{r,echo=FALSE, message=FALSE, warning=FALSE, fig.height=2, fig.width=3, fig.align='center'}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv))+
  geom_point()+
  geom_smooth(se = FALSE)

```


```{r,echo=FALSE, message=FALSE, warning=FALSE, fig.height=2, fig.width=3, fig.align='center'}
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy, color = drv))+
  geom_smooth(mapping = aes(x = displ, y = hwy),color = "blue", se = FALSE)
```


```{r,echo=FALSE, message=FALSE, warning=FALSE, fig.height=2, fig.width=3, fig.align='center'}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy))+
  geom_point(mapping = aes(color = drv))+
  geom_smooth(mapping = aes(linetype = drv), se = FALSE)
```


```{r,echo=FALSE, message=FALSE, warning=FALSE, fig.height=2, fig.width=3, fig.align='center'}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy))+
  geom_point(mapping = aes(color = drv))
  
```


###### Answers:


```{r, fig.show='hide', warning=FALSE, message=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy))+
  geom_point()+
  geom_smooth(se = FALSE,color = "blue")
```


```{r, fig.show='hide', warning=FALSE, message=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy))+
  geom_point()+
geom_smooth(se = FALSE, mapping = aes(x = displ, y = hwy, group_by =drv), show.legend = FALSE)

```


```{r, fig.show='hide', warning=FALSE, message=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv))+
  geom_point()+
  geom_smooth(se = FALSE)

```


```{r,fig.show='hide', warning=FALSE, message=FALSE}
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy, color = drv))+
  geom_smooth(mapping = aes(x = displ, y = hwy),color = "blue", se = FALSE)
```


```{r,fig.show='hide', warning=FALSE, message=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy))+
  geom_point(mapping = aes(color = drv))+
  geom_smooth(mapping = aes(linetype = drv), se = FALSE)
```


```{r,fig.show='hide', warning=FALSE, message=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy))+
  geom_point(mapping = aes(color = drv))
  
```


***Question: What is the default geom associated with ``stat_summary``?***

###### Answer:

``geom_point``


***Question: What does ``geom_col()`` do? How is it different from ``geom_bar()?``***

###### Answer: 

**``geom_col`` represents the heights of the bars with the values in the data set while geom_bar() makes the height of the bar proportional to the number of cases( that is occurrences) in each group**

***Question: What variables do stat_smooth() compute?***

##### Answer:
**The predicted value, lower value of confidence interval and upper value confidence interval**

***Question: What is the problem with this plot? How could you improve it?***
```{r, message=FALSE, warning=FALSE, fig.height=2, fig.width=3}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy))+
  geom_point()
```

##### Answer:
**The outcome shows there is over-plotting on the visual due to the discreteness in smaller datasets. To improve this plot, the argument ``position = jitter`` should be supplied to the ``geom_point()`` to add random variations to the location of each point in the over-plotted visual as shown below**

```{r, error=FALSE, message=FALSE, fig.height=2, fig.width=3}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy))+
  geom_point(position = "jitter")

```

###### Alternative metod: **Use geom_jitter() instead of geom_point()**
```{r, error=FALSE, message=FALSE, fig.height=2, fig.width=3}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy))+
  geom_jitter()
```

###### What parameters in ``geom_jitter`` control the amount of jittering?

***Answer:***
**The ``width`` which controls the amount of horizontal displacement and the ``height`` which controls the amount of vertical displacement**

###### Compare and contrast ``geom_jitter`` and ``geom_count``

**Answer:**

* ``geom_jitter`` is a convenient variation of ``geom_point`` which adds a random variation to the locations of each points in the dataset, useful in handling over-plotting

* ``geom_count`` is a variant of ``geom_point`` that counts the number of observations at each location, then maps the count to point area.

* Both ``geom_jitter`` and ``geom_count`` declare equal aesthetics arguments 

* Both geom variants can declare ``position`` argument to override the defaults locations of datasets on the plots

##### To know more execute the commands: ``?geom_jitter`` and ``?geom_count``




