---
name: Chinedu
surname: Anyanwu
position: "Full-stack Web Developer"
address: "Owerri, Imo State Nigeria"
phone: +234-8161450553
pronouns: He/His
email: "anyanwuchinedu687@yahoo.com"
twitter: _edugreat
github: edugreat
linkedin: chinedu-anyanwu-313003236/
#date: "`r format(Sys.time(), '%B %Y')`"
aboutme: "Chinedu, is a passionate Web Developer and Data Analyst with experience in developing full stack web applications. He is adept in writing clear, concise and readable codes that are easy to understand, maintain and debug. He can be easily adaptive to changing scenarios, multitaskful, critical & analytical in thinking as well as quick in learning new technologies. He is able to work with teams or independently as the need may arise."
output: vitae::twentyseconds
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(vitae)
```

# Work Experience
```{r}
library(tibble)
tribble(
  ~ org, ~ date, ~task,
  "Personal project", "2022-date"," E-LEARNING PLATFORM"
) %>% detailed_entries(what = task, when = date, with = org)
```
\vspace{-8truemm}

 * Created e-learning platform for secondary school students, with backend data storage, Angular frontend, RESTful communication, and Git version control 
 
 \vspace{5truemm}

```{r}
library(tibble)
tribble(
  ~ org, ~ date, ~ task, ~ where,
  "UNIZIK, Awka","2017", " electronic voting system demo","Nigeria"
) %>% detailed_entries(what = task, when = date, where = where, with = org)
  
```
* Implemented logic used in a faculty-wide election process that validated, tracked and recorded voting process.

\vspace{4truemm}

# EDUCATION
```{r}
library(tibble)
tribble(
  ~ university, ~ date,~ where,~cert,
  "Nnamdi Azikiwe University", " 2015-2020", "Awka, Nigeria","B.eng (electrical & electronic engineering)"
) %>% detailed_entries(what = cert, with =university, when = date, where = where )
```
\vspace{4truemm}

# CERTIFICATIONS & BADGES

```{r}
library(tibble)
tribble(
  ~ org, ~ date, ~task,
  "", "November 2023"," Hewlett Packard Enterprise Software Engineering Job Simulation"
) %>% detailed_entries(what = task, when = date, with = org)
```
\vspace{-6truemm}

 * Wrote a proposal for a RESTful web service to manage a list of employees
 * Built a web server application in Java Spring Boot that can accept and respond to HTTP requests as well as support uploading JSON data
 * Developed and ran a set of unit tests to assess my Java Spring Boot application’s performance.

\vspace{5truemm}



```{r}
library(tibble)
tribble(
  ~ org, ~ date, ~task,
  "virtual experience program on Forage", "November 2023"," JPMorgan Chase & Co. Software Engineering Lite"
) %>% detailed_entries(what = task, when = date, with = org)
```
\vspace{-6truemm}

 * Completed a simulation focused on the process of completing an engineering ticket for a system in the credit-card rewards department of JPMorgan Chase & Co..
 * Created a new class to get an existing system up and running
 * Wrote a test suite for the class added.

\vspace{5truemm}

```{r}
library(tibble)
tribble(
  ~ award, ~ org,~ date,
  "Data Analytics", " Google", "2023"
) %>% brief_entries(
  glue::glue("Certification in {award}"),
  with = org, when = date
)
```
\vspace{-6truemm}

 * Clean, organize, and process data for analysis using spreadsheet, SQL, and R.
 * Visualize and present data in dashboards.

\vspace{5truemm}

```{r}
library(tibble)
tribble(
  ~ award, ~ org,~ date,
  "developer",
  " Udemy","2022"
) %>% brief_entries(
  glue::glue("Full stack angular & spring-boot {award}"),
  with = org, when = date
)
```
\vspace{-6truemm}

* Developed a Full Stack E-commerce app with Angular front-end and Spring Boot back-end, including a shopping cart and Stripe payment integration

\vspace{5truemm}
```{r}
library(tibble)
tribble(
  ~ award, ~ org,~ date,
  "developer"," Udemy","2021"
) %>% brief_entries(
  glue::glue("Spring & Hibernate {award}"),
  with = org, when = date
)
```
* Key Spring Boot core features including: Annotations, Java Config, Spring REST, Spring MVC, AOP, Hibernate/JPA, and Maven.
* Proficiency in Spring Boot REST APIs, Spring Boot Security, Spring Data JPA, Spring Data REST, and Thymeleaf.

# PROGRAMMING SKILLS
* Proficient in Java, JavaScript, TypeScript, SQL, and RESTful web applications using Spring Boot and Angular.
* Strong knowledge of HTML, CSS, and Bootstrap.

# SOFT SKILLS
* Good listener, adaptable, fast learner, effective English communication and writing skills, and excellent time management.


# Hobbies
* Reading books and documentaries.
* Listening to soft tones(at leisure times)

# Reference
Available on request.

